---
title: "G0O02a: Statistical Data Analysis: Project 1 - 2025/2026 Spring"
author: "Gabriela Rybacka"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
library(car)
library(robustbase)
library(ggplot2)
library(moments)
library(corrplot)
library(rrcov)
```

# Loading Data

```{r}
set.seed(1125191)

sweetsdata <- read.csv("sweets_data.csv")

X <- as.matrix(sweetsdata[ , -9])
y <- factor(sweetsdata$labels, levels = c("biscuits", "cakes"))
mygroup_ind <- which(rmultinom(1, 1, rep(1/2,2)) == 1)
mygroup <- levels(y)[mygroup_ind]
mysample <- sample(which(y == mygroup), 200)
mydatafull <- data.frame(X[mysample, ])
mytrainingdata <- mydatafull[1:125, ]
myvalidationdata <- mydatafull[126:200, ]
```

# Questions


## Question 1

```{r}
str(mydatafull)
summary(mydatafull)
head(mydatafull)
```


### Histograms

```{r}
par(mfrow = c(2, 4))

hist(mydatafull$energy_kcal, main = "Histogram of energy_kcal")
hist(mydatafull$water, main = "Histogram of water")
hist(mydatafull$carbohydrates_available, main = "Histogram of carbohydrates_available")
hist(mydatafull$sugar, main = "Histogram of sugar")
hist(mydatafull$dietary_fibres, main = "Histogram of dietary_fibres")
hist(mydatafull$fat_total, main = "Histogram of fat_total")
hist(mydatafull$fatty_acids_saturated, main = "Histogram of fatty_acids_saturated")
hist(mydatafull$Na, main = "Histogram of Na")

par(mfrow = c(1, 1))

```


### Discussion:

By looking at the histograms, we can clearly see that most of the variables have rightward skewed distribution, which means that there are more values concentrated on the left side of the distribution and fewer values on the right side.

This is seen in the variables such as "water", "dietary_fibres", "Na", and also in "fatty_acids_saturated". 

The variable "energy_kcal", "carbohydrates_available", and "sugar" appear to have a more normal distribution, although it's not perfectly normal. 

The presence of skewness in the data can affect statistical analyses and may require transformation. 


### QQ-Plots

```{r}
par(mfrow = c(2, 4))

qqnorm(mydatafull$energy_kcal, main = "QQ Plot of energy_kcal")
qqnorm(mydatafull$water, main = "QQ Plot of water")
qqnorm(mydatafull$carbohydrates_available, main = "QQ Plot of carbohydrates_available")
qqnorm(mydatafull$sugar, main = "QQ Plot of sugar")
qqnorm(mydatafull$dietary_fibres, main = "QQ Plot of dietary_fibres")
qqnorm(mydatafull$fat_total, main = "QQ Plot of fat_total")
qqnorm(mydatafull$fatty_acids_saturated, main = "QQ Plot of fatty_acids_saturated")
qqnorm(mydatafull$Na, main = "QQ Plot of Na")

par(mfrow = c(1, 1))

```


### Discussion:

The QQ plots confirm my observations from the histograms. 

The points in the QQ plots for "energy_kcal", "carbohydrates_available", and "sugar" are closer to the straight line, suggesting that these variables may follow a normal distribution. 

However, the points for "water", "dietary_fibres", "fat_total", "fatty_acids_saturated", and "Na" deviate significantly from the straight line, especially in the tails, indicating that these variables do not follow a normal distribution and are likely skewed. 

This further supports my earlier conclusion about the skewness of these variables.

### Scatter Plots

```{r}
pairs(mydatafull, main = "Scatterplots for all variables")
```


### Discussion:

There is a strong positive correlation between energy_kcal and fat_total.

Energy_kcal and fatty_acids_saturated also show a positive correlation, but it is not as strong.

Similarly, fat_total and fatty_acids_saturated have a positive correlation, though it is weaker than the one between energy and fat.

There is a negative correlation between carbohydrates_available and fat_total.


## Question 2


### Shapiro-Wilk Test for Normality

```{r}
shapiro.test(mydatafull$sugars)
shapiro.test(mydatafull$dietary_fibres)
```


### Discussion:

**H0:** Variables are normally distributed.

**H1:** Variables are not normally distributed.

The Shapiro-Wilk test results indicate that both "sugars" and "dietary_fibres" do not follow a normal distribution, as the p-values are less than 0.05.

I need to preform a transformation.

To avoid data leakage, I did the Box-Cox transformation on the training set.

I will use the lambda to transform both datasets later.


### Sugars

```{r}
shapiro.test(mytrainingdata$sugars)
skewness(mytrainingdata$sugars)
```


### Discussion:

The Shapiro-Wilk test on the training set shows a p-value > 0.05 so there is no evidence to reject the null hypothesis. 
Although the full dataset shows a different result, I decided to proceed without transforming this variable because the training data satisfies the normality assumption.


### Dietary fibres

```{r}
shapiro.test(mytrainingdata$dietary_fibres)
skewness(mytrainingdata$dietary_fibres)

sum(is.na(mytrainingdata$dietary_fibres))
sum(mytrainingdata$dietary_fibres == 0)

shapiro.test(sqrt(mytrainingdata$dietary_fibres +1))
skewness(sqrt(mytrainingdata$dietary_fibres +1))

shapiro.test(1/(mytrainingdata$dietary_fibres +1))
skewness(1/(mytrainingdata$dietary_fibres +1))

shapiro.test(log(mytrainingdata$dietary_fibres +1))
skewness(log(mytrainingdata$dietary_fibres +1))

shapiro.test(1/(mytrainingdata$dietary_fibres +1)^2)
skewness(1/(mytrainingdata$dietary_fibres +1)^2)

tempdf <- powerTransform(mytrainingdata$dietary_fibres + 1)
lambda_fibres <- tempdf$lambda
xdf <- mytrainingdata$dietary_fibres
x.primdf <- bcPower(mytrainingdata$dietary_fibres + 1, lambda_fibres)
shapiro.test(x.primdf)
skewness(x.primdf)
```


### Discussion:

The distribution of dietary fibres is highly skewed (2.32) and contains zero values (2). Its distribution also is not normal. 

To fix this (the 0 values), I added 1 to all observations before testing transformations. 

Simple methods like square root or log improved the skewness but failed the Shapiro-Wilk test. 

The Box-Cox (+1) transformation was the most effective because it reduced the skewness to nearly zero (-0.04). Although the p-value (0.001) was still below 0.05, this transformation is the best possible option to bring the data closer to a normal distribution.


## Data transformation

```{r}
mytrainingdata$dietary_fibres <- bcPower(mytrainingdata$dietary_fibres + 1, lambda_fibres)
myvalidationdata$dietary_fibres <- bcPower(myvalidationdata$dietary_fibres + 1, lambda_fibres)

par(mfrow = c(1, 2))

hist(mydatafull$dietary_fibres, main = "Histogram of dietary_fibres")
hist(x.primdf, main = "Histogram of dietary_fibres after Box-Cox transformation", col = "lightblue")

par(mfrow = c(1, 1))
```


## Question 3


### Scatter Plot - Elipses

```{r}
X <- mytrainingdata[, c("carbohydrates_available", "fat_total")]

(mcd25 <- covMcd(X, alpha = 0.75))
(mcd50 <- covMcd(X, alpha = 0.5))

m <- as.numeric(colMeans(X))
S<- as.matrix(cov(X))
c <- sqrt(qchisq(0.99, df = 2))


plot(X, pch = 20, xlim=c(30,100), ylim=c(-5,55), col = "grey")


car::ellipse(center = m, shape = S, radius = c, 
             col = "brown", lty = 1, add = TRUE)

car::ellipse(center = as.numeric(mcd50$center), shape = mcd50$cov, radius = c, 
             col = "lightblue", lty = 2, add = TRUE)

car::ellipse(center = as.numeric(mcd25$center), shape = mcd25$cov, radius = c, 
             col = "violet", lty = 3, add = TRUE)

legend("topright", legend = c("classical", "MCD 50%", "MCD 25%"),
       col = c("brown", "lightblue", "violet"), lty = 1:3, lwd = 2)
```


### Discussion:

The classical tolerance ellipse (brown) is significantly larger and more elongated than the robust MCD ellipses. 

This indicates that standard covariance and mean can be inflated by outliers, failing to represent the true form of the data.


### Comparison of the robust distances with the Mahalanobis distances

```{r}
md <- sqrt(mahalanobis(X, center = m, cov = S))
rd <- sqrt(mahalanobis(X, center = mcd50$center, cov = mcd50$cov))

kolory_punktow <- rep("grey60", length(md))
kolory_punktow[md <= c & rd > c] <- "lightblue"
kolory_punktow[md > c & rd > c] <- "red"

plot(md, rd, 
     pch = 20, 
     col = kolory_punktow,
     main = "DD plot",
     xlab = "Mahalanobis distance",
     ylab = "MCD distance")

abline(h = c, lty = 2, col = "red")
abline(v = c, lty = 2, col = "red")

legend("topleft", 
       legend = c("normal", "MCD only", "outliers"),
       col = c("grey60", "lightblue", "red"), 
       pch = 20, cex = 0.9)

sum(kolory_punktow == "lightblue")
sum(kolory_punktow == "red")
```


### Discussion:

The DD-plot reveals 18 outliers (MCD only). 

These observations were classified as "normal" by the standard Mahalanobis distance but are correctly flagged as anomalies by the robust MCD distance.

The DD-plot also reveals 1 observation that is an outlier according to both measures.


## Question 4


### Part a

```{r}
boxplot(mytrainingdata)

summary(mytrainingdata)
```


### Discussion:

I need to standardize the data before performing PCA because the variables are on different scales.

Especially we can see the difference between enery_kcal and the rest ones.

Moreover PCA is sensitive to the scale of the data, and standardizing ensures that each variable contributes equally to the analysis.

Especially after the box-cox transformation, the scales of the variables might have changed, making standardization even more needed.


### Part b

I am going to choose the number of principal components by using the scree plot and the cumulative variance values.

```{r}
pca_model <- prcomp(mytrainingdata, scale = TRUE)

plot(pca_model, type = "lines", main = "Scree Plot")
abline(h = 1, col = "red", lty = 2)

summary(pca_model)

pca_model <- PcaClassic(mytrainingdata, k = 4, scale = TRUE, crit.pca.distances = 0.99)
```


### Discussion:

I selected four principal components basing on the scree plot and the Kaiser criterium. 

Only the first four components have a variance greater than 1. 

The analysis of cumulative variance shows that the first 4 components collectively explain 89.96% of the total variance in the data. 

This makes the data much simpler while keeping almost all the important information.


### Part c

```{r}
scaled_data <- scale(mytrainingdata)
boxplot(scaled_data, main = "Boxplot after standardization", col = "lightblue")

(cor_matrix <- cor(mytrainingdata, pca_model$scores))
round(cor_matrix[, 1:4], 2)

corrplot(cor_matrix[, 1:4], method = "color", type = "full", 
         addCoef.col = "black",
         tl.col = "black", tl.srt = 45,
         number.cex = 0.7)
```


### Discussion:

The first principal component (PC1) shows the difference between fat_total and energy_kcal vs. carbohydrates.

It has a strong positive correlation with fat_total (0.98), energy_kcal (0.92), and fatty_acids_saturated (0.81), and a strong negative correlation with carbohydrates_available (-0.73). 

The second component (PC2) shows the difference between water and carbohydrates_available and Na in sweets.

It is positively correlated with water (0.76) and negatively correlated with Na (-0.69) and carbohydrates (-0.58). 

The third component (PC3) has a strong positive correlation with sugars (0.74) and a negative correlation with dietary fiber (-0.64).

The fourth component (PC4) has a positive correlation with dietary_fibres(0.62) and a negative correlation with water (-0.55).


### Part d


### Discussion:
 
   
#### **PC1 vs PC2**

```{r}
rrcov::biplot(pca_model, choices = c(1, 2), scale = 0)
```

On the first biplot, the red arrows for fat and energy point to the right along the PC1 axis, while the arrow for carbohydrates points to the left. 

For example, observation number 116 is located far on the right side, meaning it is a very fatty and high-calorie sweet.

On the PC2 axis, the water arrow points up, while Na and carbs point down. Observation 1 is at the very top, showing it is a sweet with a lot of water, while observations like 72, 89 and 91 are in the bottom left area, meaning they are rich in carbs (not as much 91) and Na but low in fat.


#### **PC1 vs PC3**

```{r}
rrcov::biplot(pca_model, choices = c(1, 3), scale = 0)
```

In the PC1 vs PC3 biplot, the arrow for sugars points almost straight up, and the fiber arrow points down. 

For example observation number 80, which sits very high on the PC3 axis, has a very high sugar content.

Observation 116 stands out on the far right side, meaning it has a very high fat and calories content.
Observation 110 is positioned far to the left, which means that this sweet is very high in carbohydrates but low in fat.


#### **PC2 vs PC3**

```{r}
rrcov::biplot(pca_model, choices = c(2, 3), scale = 0)
```

In the PC2 vs PC3 biplot, observations 72 and 91 stand out as sweets with very high Na and carbohydrate levels.
Observation 80 is at the very top, making it the most extreme outlier for sugar content.


## Question 5


### Part a

```{r}
pca_class <- PcaClassic(mytrainingdata, k = 4, scale = TRUE, crit.pca.distances = 0.99)
pca_rob <- PcaHubert(mytrainingdata, k = 4, scale = mad, crit.pca.distances = 0.99)

val_scores <- predict(pca_rob, newdata = myvalidationdata)

val_fitted <- t(t(val_scores %*% t(pca_rob@loadings)) + pca_rob@center / pca_rob@scale)
val_scaled_no_center <- scale(myvalidationdata, center = FALSE, scale = pca_rob@scale)

euclnorm <- function(y) sqrt(sum(y^2))
val_od <- apply(val_scaled_no_center - val_fitted, 1, euclnorm)
val_sd <- sqrt(mahalanobis(val_scores, center = FALSE, cov = diag(pca_rob@eigenvalues)))

cutoff.sd <- pca_rob@cutoff.sd
cutoff.od <- pca_rob@cutoff.od

ods <- c(pca_rob@od, val_od)
sds <- c(pca_rob@sd, val_sd)

xmax <- max(sds, cutoff.sd, na.rm = TRUE)
ymax <- max(ods, cutoff.od, na.rm = TRUE)

colors <- c(rep("lightblue", nrow(mytrainingdata)), rep("red", nrow(myvalidationdata)))
pchs <- c(rep(16, nrow(mytrainingdata)), rep(17, nrow(myvalidationdata)))

plot(sds, ods, pch = pchs, col = colors,
     xlim = c(0, xmax + 1), ylim = c(0, ymax + 0.2),
     main = "Robust PCA: Outlier Map",
     xlab = "Score distance", ylab = "Orthogonal distance")

abline(h = cutoff.od, col = "red", lwd = 1.5)
abline(v = cutoff.sd, col = "red", lwd = 1.5)

legend("topright", legend = c("Training Set", "Validation Set"),
       col = c("lightblue", "red"), pch = c(16, 17))
```


### Discussion:

The outlier map shows Score Distance on the horizontal axis and Orthogonal Distance on the vertical axis. 

The red lines are the safety limits. Any point outside these lines is an outlier. 
We can see many blue circles (training set) and red triangles (validation set) that cross such limits. 
These points would ruin a classical PCA analysis. In a classical model, they would pull the results toward themselves and distort the true patterns of our data.


### Part b
```{r}
bad_leverage_indices <- which(pca_rob@od > cutoff.od & pca_rob@sd > cutoff.sd)

mytrainingdata[bad_leverage_indices, c("carbohydrates_available", "sugars", "fat_total", "Na")]
```


### Discussion:

These points are flagged as outliers because their nutritional values, like extremely high sugar or fat, are very unusual compared to the rest sweets.


### Part c

```{r}
par(mfrow = c(1, 2))

col_class <- ifelse(pca_class@sd > pca_class@cutoff.sd | pca_class@od > pca_class@cutoff.od, "red", "black")

plot(pca_class@sd, pca_class@od, pch = 19, col = col_class,
     xlab = "Score distance", ylab = "Orthogonal distance", 
     main = "Classical PCA")
abline(v = pca_class@cutoff.sd, h = pca_class@cutoff.od, col = "red", lty = 2)

outliers_class <- which(col_class == "red")
text(pca_class@sd[outliers_class], pca_class@od[outliers_class], 
     labels = outliers_class, pos = 3, cex = 0.7, col = "red")

col_rob <- ifelse(pca_rob@sd > pca_rob@cutoff.sd | pca_rob@od > pca_rob@cutoff.od, "red", "black")

plot(pca_rob@sd, pca_rob@od, pch = 19, col = col_rob,
     xlab = "Score distance", ylab = "Orthogonal distance", 
     main = "Robust PCA")
abline(v = pca_rob@cutoff.sd, h = pca_rob@cutoff.od, col = "red", lty = 2)

outliers_rob <- which(col_rob == "red")
text(pca_rob@sd[outliers_rob], pca_rob@od[outliers_rob], 
     labels = outliers_rob, pos = 3, cex = 0.7, col = "red")

par(mfrow = c(1, 1))
```


### Discussion:

The comparison shows a big difference between the two methods. 

Classical PCA tries to include every single point, even the extreme ones. 

Because of this, it hides the real outliers by making their distances look smaller. 

For example, observation 80 does not look very extreme on the classical plot. 

However, robust PCA ignores these bad points. 

As a result, the robust map clearly reveals the true outliers, like 80, 65, 74, and 107, showing them with much higher distances. 

This proves that robust PCA is much better and more accurate for this data.